Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-03-22 paper/paper.tex
Some corrections

\begin{abstract}
In this work we propose a Bayesian approach for functional linear or logistic  regression models, based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional alternative to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subset of the RKHS generated by the underlying covariance function. By imposing a suitable prior distribution on such RKHS, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be approximated from Markov Chain Monte Carlo (MCMC) methods. Several estimators derived from this posterior distribution turn out to be competitive against other usual  alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection method.
\end{abstract}

@@

@@

The problem of inferring a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle we could simply treat the functional data as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably treat functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on continuous data. A recent survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} and \citet{horváth2012inference}.

In this work we are concerned with (supervised) functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some hidden dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). We will assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

The most common functional linear regression model is the classical \(L^2\)-model, originally introduced in the first edition of the book by~\citet{ramsay2005functional} \incomment{fue realmente así?}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \beta_0 + \dotprod{X}{\beta} + \epsilon = \beta_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
\incomment{mejor poner el producto con el subíndice \(\langle \cdot , \cdot \rangle_2\)?}~where \(\beta_0\in \R\) and \(\epsilon\) is an independent error term with certain simplifying assumptions. In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinite-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle drawbacks that arise in this \(L^2\)-framework. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-linear-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case \citep[see][]{berrendero2019rkhs}. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\beta_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\beta_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this case, the most common way of estimating the slope function \(\beta(t)\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{buenolarraz2021functional}.

It turns out that in both situations a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. These RKHS-based models have already been explored in \citet{berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018use} for the case of functional classification. The main contribution of this work is a parametric Bayesian approach to parameter estimation within these recently-proposed models, in which a prior distribution is imposed on the unknown function to  obtain a posterior distribution after seeing the data.

Another set of techniques widely studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_i)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we could easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_i\}\), and thus provide a variable selection method along with the other estimators that naturally arise within our model. In this way, in addition to making predictions about the input data, we could evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also appeared recently in the literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration.

\begin{comment}
  Quizás falta comentar algo de los resultados experimentales... también habría que hacer más hincapié en cuál es nuestra propuesta; que se vea bien.
\end{comment}

@@

\begin{comment}
¿Mejor poner esto después de la sección 1.1?
\end{comment}

In Section~\ref{sec:rkhs} we carry out a quick review of the theory of RKHS's from a probabilistic point of view. Section~\ref{sec:methodology} is devoted to explaining the Bayesian methodology and the functional regression models proposed. The empirical results of the experimentation are contained in Section~\ref{sec:results}, which also includes a short discussion of computational details. Lastly, the conclusions drawn from this work are presented in Section~\ref{sec:conclusion}.

\subsection{A brief review of the theory of RKHS's}\label{sec:rkhs}

\begin{comment}
  Es posible que esta sección sea demasiado larga...?
\end{comment}

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces (for a more detailed account, see for example~\citet{berlinet2004reproducing}). In what follows, suppose \(X=X(t)\) is an \(L^2\)-process with trajectories in \(L^2[0,1]\), and further assume that its mean function \(m(t)=\mathbb E[X(t)]\) and its covariance function \(K(t, s)= \mathbb E[(X(t) - m(t))(X(s) - m(s))]\) are both continuous. To construct the RKHS associated with the covariance function, we start by defining the functional space \(\mathcal H_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\[
\mathcal H_0(K) = \left\{ f \in L^2[0,1]: f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\]
This space is endowed with the inner product
\[
\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j),
\]
given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Then, we define \(\mathcal H(K)\) to be the completion of \(\mathcal H_0(K)\) under the norm induced by the scalar product defined above. As it turns out, functions in this space verify the \textit{reproducing property}, thus giving it its name:
\begin{equation}\label{eq:reproducing-property}
  \dotprod{K(t, \cdot)}{f}_K = f(t), \quad \text{for all } f \in \mathcal H(K), \ t \in [0, 1].
\end{equation}
An important detail to note is that \(\mathcal H(K)\) is a space of genuine functions, and not of equivalence classes of functions. Hence, the values of these functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

As we can see, the covariance function (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS \incomment{¿Es necesario destacar alguna suposición/propiedad extra sobre \(K\)? Por ejemplo \(\int\int K^2 <\infty\) o que es simétrica y (semi-) definida positiva}. A linear operator \incomment{No sé muy bien qué calificativos ponerle: integral Hilbert-Schmidt operator, compact, self-adjoint, bounded, linear functional, ...}\ closely related to this covariance function is the so-called covariance operator, namely
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
This operator is especially relevant due to the \textit{Karhunen-Loève} theorem~\cite[e.g.][Th.~1.5]{bosq2000linear}, which asserts that a centered process \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{i=1}^\infty \xi_i \phi_i(t),
\end{equation}
where \(\phi_i\) are the (orthonormal) eigenfunctions of its covariance operator \(\mathcal K\) and \(\xi_i = \int X\phi_i\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_i\), the eigenvalue associated with \(\phi_i\). This expansion is the main justification for many FDA techniques, such as functional principal component analysis (FPCA)~\cite[see for example][Ch.~8]{ramsay2005functional}.

@@

\begin{description}
  \item[From an isometry] Via \textit{Loève's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) (see Lemma 1.1 in \citet{lukic2001stochastic}). This isometry is essentially the completion of the correspondence
  \[
  \sum_{i=1}^p a_i (X(t_i) - m(t_i)) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
\]
and can be formally defined as
\begin{equation}\label{eq:loeves-isometry}
  \Psi_X(U)(t) = \E[U(X(t) - m(t))], \quad U \in \mathcal L(X).
\end{equation}
  \item[From the covariance operator] The space \(\mathcal H(K)\) can also be identified with the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}
In this case, the inner product can be expressed as
\[
\dotprod{f}{g}_K = \dotprod{\mathcal K^{1/2}(f)}{\mathcal K^{1/2}(g)}.
\]

@@

\begin{comment}
  Puede que las definiciones 2ª y 3ª arriba sean innecesarias.
\end{comment}

@@

In this section we present the precise models and Bayesian methodologies explored in this work. In the sequel, we will assume the notation from the previous sections and suppose without loss of generality that \(X\) is centered, i.e., \(m(t)=0\) for all \(t\in[0, 1]\). The RKHS-based functional models under consideration are those obtained by taking a functional parameter \(\alpha \in \mathcal H(K)\) and replacing the scalar product for \(\dotprod{x}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subset \(\mathcal H_0(K)\), i.e.,
\begin{equation}\label{eq:alpha-parameter-rkhs}
  \alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot) \in \mathcal H_0(K).
\end{equation}

@@

The general idea will be to impose a prior distribution on these free parameters (and possibly others) to eventually derive a posterior model after incorporating the evidence of the available sample data. In view of expression~\eqref{eq:alpha-parameter-rkhs}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\mathcal H_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\{\beta_j\}\) and the times \(\{t_j\}\) separately. Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x^{-1}(\alpha)\), where \(\Psi\) is Loève's isometry. Hence, taking into account that \(\Psi_X^{-1}(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we have
\[
\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j).
\]

On a separate note, it is worth mentioning that starting from a probability distribution \(\mathbb{P}_0\) on \(\mathcal H_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\mathcal H(K)\) simply by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \mathcal H_0(K))\) for all Borel sets \(B \in \mathcal B(\mathcal H(K))\). This is why our simplifying assumption on \(\alpha\) is not actually very restrictive, since any prior distribution on \(\mathcal H_0(K)\) can be directly extended to a prior distribution on \(\mathcal H(K)\). The following result proves that any such extension is indeed unique.

\begin{prop} Let \(H\) be a separable Hilbert space and consider a dense subset \(H_0\subseteq H\). Then, any probability distribution on \((H, \mathcal{B}(H))\) is uniquely determined by its restriction to \((H_0, \mathcal B(H_0))\).
\end{prop}

@@

@@

In much the same way as the functional linear regression model explained in the previous section, this RKHS-based logistic regression model partially solves some of the issues with the \(L^2\)-model~\eqref{eq:l2-logistic-model} stated in Section~\ref{sec:intro}. Moreover, it negates the problem of non-existence of the MLE in the functional case, since our proposed model is finite-dimensional\footnote{Even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist. However, this problem can be easily circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.}. Additionally, under specific conditions this model also entails the logistic \(L^2\)-model in the Gaussian homoscedastic case \citep[see][Th.~1]{buenolarraz2021functional}.

>>>>

\begin{abstract}
We propose a Bayesian approach for functional linear or logistic  regression models, based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subset of the RKHS generated by the underlying covariance function. By imposing a suitable prior distribution on such RKHS, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be approximated from Markov Chain Monte Carlo (MCMC) methods. Several estimators derived from this posterior distribution turn out to be competitive against other usual  alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection method.
\end{abstract}

@@

\begin{comment}
  \begin{itemize}
    \item Utilizar \(\alpha\) como parámetro en los modelos \(L^2\)?
    \item Qué se demuestra exactamente en la proposición 2.1?
    \item Suponer que los datos \(X\) están centrados desde el principio.
    \item Definir la isometría de Loève al revés
  \end{itemize}
\end{comment}

@@

The problem of predicting a scalar response from a functional covariate is one that has gained traction over the last few decades, as more and more data is being generated with an ever-increasing level of granularity in the measurements. While in principle we could simply regard the functional data as a discretized vector in a very high dimension, there are often many advantages in taking into account the functional nature of the data, ranging from modeling the possibly high correlation among points that are close in the domain, to extracting information that may be hidden in the derivatives of the function in question. As a consequence, numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on continuous data. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} and \citet{horváth2012inference}.

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some more subtle drawbacks that emerge as a result of working in infinite dimensions. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\). In addition, without loss of generality we will suppose that \(X\) is centered, i.e., that its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). Furthermore, we will assume the existence of a ``labeled'' data set \(\mathcal D_n =\{(X_i, Y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), and our aim will be to accurately predict the response corresponding to unlabeled samples from \(X\).

\subsubsection{\(L^2\)-models, shortcomings and alternatives}

 The most common functional linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the book by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^d\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is an error term independent from \(X\) with \(\E [\epsilon]=0\), and the functional slope parameter \(\beta=\beta(\cdot)\) is supposed to be a member of the infinite-dimensional space \(L^2[0, 1]\). In this case, the inference on \(\beta\) is hampered by the fact that \(L^2[0,1]\) is an extremely wide space that contains many non-smooth or ill-behaved functions, so that any estimation procedure involving optimization on it would typically be very hard. In spite of this, model~\eqref{eq:l2-linear-model} is not flexible enough to include ``simple'' finite-dimensional models based on a linear combination of the marginals, such as \(Y=\beta_0 + \beta_1 X(t_1)+ \cdots + \beta_p X(t_p) + \epsilon\) for some constants \(\beta_j\in\R\) and instants \(t_j\in[0,1]\) \citep[see][]{berrendero2019rkhs, berrendero2020general}. Moreover, the non-invertibility of the covariance operator associated with \(X\) (defined in Section~\ref{sec:rkhs}), which plays the role of the covariance matrix in the infinite case, invalidates the usual least squares theory.

A similar \(L^2\)-based functional logistic equation can be derived for the binary classification problem via the logistic function:
\begin{equation}\label{eq:l2-logistic-model}
  \mathbb P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\beta}\}},
\end{equation}
where \(\alpha_0 \in \R\) and \(\beta \in L^2[0, 1]\). In this situation, the most common way of estimating the slope function \(\beta(\cdot)\) is via its Maximum Likelihood Estimator (MLE). However, not only do the same complications as in the linear regression model apply in this situation, but there is also the additional problem that in functional settings the MLE does not exist with probability one under fairly general conditions \citep[see][Sec.~3.2]{berrendero2021functional}.

It turns out that in both situations a natural alternative to the \(L^2\)-model is the so-called Reproducing Kernel Hilbert Space (RKHS) model, which instead assumes the unknown parameter to be a member of the RKHS associated with the covariance function of the process \(X\), making use of the scalar product of that space. As we will show later on, not only is this model simpler and arguably easier to interpret, but it also constrains the parameter space to smoother and more manageable functions. In fact, it includes a model based on finite linear combinations of the marginals of \(X\) as a particular case, which is particularly appealing to practitioners confronted with functional data problems due to its simplicity. These RKHS-based models and their idiosyncrasies have been explored in \citet{berrendero2019rkhs, berrendero2020general} in the functional linear regression setting, and in \citet{berrendero2018use, berrendero2021functional} for the case of logistic regression.

A major aim of this work is to motivate these recently-proposed models inside the functional framework, while also providing efficient techniques to apply them in practice. Our main contribution is the proposal of a Bayesian approach to parameter estimation within the aforementioned RKHS models, in which a prior distribution is imposed on the unknown functional parameter to obtain a posterior distribution after seeing the data. In this way, one can incorporate pre-existing beliefs about the parameters, thus increasing the flexibility of the estimation procedure. Even though setting a prior distribution on a function space is generally a hard task, the specific parametric formulation of the RKHS models we propose greatly facilitates this (see Section~\ref{sec:methodology} for details). \incomment{Comentar resultados experimentales}

Another set of techniques extensively studied in this context are variable selection methods, which aim to select the marginals \(\{X(t_i)\}\) of the process that better summarize it according to some optimality criterion. As it happens, some variable selection methods have already been proposed in the RKHS framework \citep[see for example][]{berrendero2019rkhs}, but in general they have their own dedicated algorithms and procedures. As will become apparent in the forthcoming sections, given the nature of our suggested Bayesian model we could easily isolate the marginal posterior distribution corresponding to a finite set of points \(\{t_i\}\), and thus provide a Bayesian-motivated variable selection process along with the other prediction methods that naturally arise within our model. In this way, in addition to making predictions about the input data, we could evaluate exactly which marginals of the functional explanatory variable contain the most relevant information. These points-of-impact selection models for functional predictors have also appeared recently in the literature; see \citet{poss2020superconsistent}, \citet{berrendero2016variable} or \citet{ferraty2010most} by way of illustration. \incomment{Comentar resultados experimentales}

@@

In Section~\ref{sec:rkhs} we review the basics of the theory of RKHS's from a probabilistic point of view. Section~\ref{sec:methodology} is devoted to explaining the Bayesian methodology and the functional regression models we propose. The empirical results of the experimentation are contained in Section~\ref{sec:results}, which also includes a short discussion of computational details. Lastly, the conclusions drawn from this work are presented in Section~\ref{sec:conclusion}.

\subsection{Some essentials on RKHS's and notation}\label{sec:rkhs}

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces (for a more detailed account, see for example~\citet{berlinet2004reproducing}). In what follows, suppose \(X=X(t)\) is a centered \(L^2\)-process with trajectories in \(L^2[0,1]\), and further assume that its covariance function \(K(t, s)= \mathbb E[(X(t) - m(t))(X(s) - m(s))]\) is continuous. To construct the RKHS associated with the covariance function, we start by defining the functional space \(\mathcal H_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\begin{equation}\label{eq:h0}
\mathcal H_0(K) = \left\{ f \in L^2[0,1]: f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\end{equation}
This space is endowed with the inner product
\[
\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j),
\]
given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Then, we define \(\mathcal H(K)\) to be the completion of \(\mathcal H_0(K)\) under the norm induced by the scalar product defined above.
\incomment{Contar definiciones alternativas (con referencias), y definir el operador lineal \(\mathcal K\) inline, diciendo que al ser K continua es compacto y autoadjunto.} As it turns out, functions in this space verify the \textit{reproducing property}, thus giving it its name:
\begin{equation}\label{eq:reproducing-property}
  \dotprod{K(t, \cdot)}{f}_K = f(t), \quad \text{for all } f \in \mathcal H(K), \ t \in [0, 1].
\end{equation}
An important detail to note is that \(\mathcal H(K)\) is a space of genuine functions, and not of equivalence classes of functions. Hence, the values of these functions at particular points are in fact relevant, unlike in \(L^2\)-spaces.

\begin{comment}
  Un enfoque particularmente útil en estadística es la isometría de Loeve....

  SUPONER QUE X ESTÁ CENTRADA DESDE EL PRINCIPIO.
\end{comment}

As we can see, the covariance function (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS. A compact self-adjoint operator \incomment{No sé muy bien qué calificativos ponerle: integral Hilbert-Schmidt operator, compact, self-adjoint, bounded, linear functional, ...}\ closely related to this covariance function is the so-called covariance operator, namely
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
\begin{comment}
  QUITAR K-L Y HABLAR DE LOS AUTOVALORES EN LA DEFINICIÓN ALTERNATIVA.
\end{comment}

This operator is especially relevant due to the \textit{Karhunen-Loève} theorem~\cite[e.g.][Th.~1.5]{bosq2000linear}, which asserts that \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{i=1}^\infty \xi_i \phi_i(t),
\end{equation}
where \(\phi_i\) are the (orthonormal) eigenfunctions of its covariance operator \(\mathcal K\) and \(\xi_i = \int X\phi_i\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_i\), the eigenvalue associated with \(\phi_i\). This expansion is the main justification for many FDA techniques, such as functional principal component analysis (FPCA)~\cite[see for example][Ch.~8]{ramsay2005functional}.

@@

\begin{description}
  \item[From an isometry] Via \textit{Loève's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) (see Lemma 1.1 in \citet{lukic2001stochastic}). This isometry is essentially the completion of the correspondence
  \[
  \sum_{i=1}^p a_i X(t_i) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
\]
and can be formally defined as \incomment{Defnir al revés y usar \(\Psi\) directamente en lo sucesivo.}
\begin{equation}\label{eq:loeves-isometry}
  \Psi_X(U)(t) = \E[U X(t)], \quad U \in \mathcal L(X).
\end{equation}
  \item[From the covariance operator] The space \(\mathcal H(K)\) can also be identified with the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}
In this case, the inner product can be expressed as
\[
\dotprod{f}{g}_K = \dotprod{\mathcal K^{1/2}(f)}{\mathcal K^{1/2}(g)}.
\]

@@

\begin{comment}
  Enlazar con la definición de isometría de loeve
\end{comment}

@@

In this section we present the precise models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration are those obtained by taking a functional parameter \(\alpha \in \mathcal H(K)\) and replacing the scalar product for Loève's isometry \(\dotprod{x}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subset \(\mathcal H_0(K)\) defined in \eqref{eq:h0}, i.e.,
\begin{equation}\label{eq:alpha-parameter-rkhs}
  \alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot) \in \mathcal H_0(K).
\end{equation}

@@

The general idea will be to impose a prior distribution on these free parameters (and possibly others) to eventually derive a posterior model after incorporating the information of the available sample data. \incomment{Una de las cosas difíciles en estadística bayesiana es poner un prior en un espacio de funciones. Una de las ventajas del modelopropuesto (y por el que advocamos por este modelo) es que facilita esta tarea, de manera natural y manejable. Poner una subsubsection con: definición de una priori adecuada en nuestro RKHS framework}In view of expression~\eqref{eq:alpha-parameter-rkhs}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\mathcal H_{0,p}(K)\)) it suffices to consider \(p\)-dimensional continuous prior distributions for the coefficients \(\{\beta_j\}\) and the times \(\{t_j\}\) separately. Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x^{-1}(\alpha)\), where \(\Psi\) is Loève's isometry. Hence, taking into account that \(\Psi_X^{-1}(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we have
\[
\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j).
\]

To begin with, note that starting from a probability distribution \(\mathbb{P}_0\) on \(\mathcal H_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\mathcal H(K)\) simply by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \mathcal H_0(K))\) for all Borel sets \(B \in \mathcal B(\mathcal H(K))\). This is why our simplifying assumption on \(\alpha\) is not actually very restrictive, since any prior distribution on \(\mathcal H_0(K)\) can be directly extended to a prior distribution on \(\mathcal H(K)\). The following result proves that any such extension is indeed unique.

\begin{prop} Let \(H\) be a separable Hilbert space and consider a dense subset \(H_0\subseteq H\). \incomment{Decirlo al revés: la distribución se extiende de forma única a partir del denso.} Then, any probability distribution on \((H, \mathcal{B}(H))\) is uniquely determined by its restriction to \((H_0, \mathcal B(H_0))\).
\end{prop}

@@

\incomment{Mencionar que en realidad trabajamos en \(H_{0,p}\).}

@@

In much the same way as the functional linear regression model explained in the previous section, this RKHS-based logistic regression model partially solves some of the issues with the \(L^2\)-model~\eqref{eq:l2-logistic-model} stated in Section~\ref{sec:intro}. Moreover, it negates the problem of non-existence of the MLE in the functional case, since our proposed model is finite-dimensional\footnote{Even in finite-dimensional settings there are cases of quasi-complete separation in which the MLE does not exist. However, this problem can be easily circumvented using, for example, Firth's corrected estimator \citep{firth1993bias}.}. Additionally, under specific conditions this model also entails the logistic \(L^2\)-model in the Gaussian homoscedastic case \citep[see][Th.~1]{berrendero2021functional}.
